Making ice sheet models \\ scale properly

### 2023-01-24 talk/slides.tex
making small things better

\medskip
\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
\item relationships for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}} \hspace{20mm}$$

\vspace{-18mm}
\hfill \includegraphics[width=0.15\textwidth]{images/one-grid.png} \hspace{5mm}

\item explicit time-stepping schemata:
{\footnotesize
\begin{align*}
\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
\end{align*}
}

@@

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$
\end{tabular}
\end{frame}

@@

\begin{itemize}
\item thus \alert{explicit} time-stepping \alert{implies many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] stress-balance solver scaling $\alpha$ is important, but even optimality ($\alpha=0$) cannot fix performance
    \end{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] need $q$ model updates per year to integrate climate influences
    \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
    \item[$\circ$] parameterize cost of these solves as $O(m^{1+\beta})$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs), but in the \alert{non-local residual case}
    \begin{itemize}
    \item[$\circ$] this seems not to exist
    \item[$\circ$] the \alert{smoother} must compute an implicit-step SKE residual including surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ from a scalable Stokes solver
    \end{itemize}
\item near-optimal multilevel VI solvers exist for toy problems
    \begin{itemize}
    \item[$\circ$] Poisson equation obstacle problem (Brandt \& Cryer 1983; Gr\"aeser \& Kornhuber 2009)
    \end{itemize}
\item<2> \alert{help me}
\end{itemize}
\end{frame}

>>>>

\medskip
\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
\item for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
  $$m \sim \frac{L^2}{h^2} \hspace{20mm}$$

\vspace{-14mm}
\hfill
\begin{tikzpicture}[scale=1.5]
  \pgfmathsetmacro\fourth{1.0/4.0}
  \draw[xstep=\fourth,ystep=\fourth,black,thin] (0.0,0.0) grid (1.0,1.0);
  \draw[<->] (-0.2,0.0) -- (-0.2,1.0);
  \draw[<->] (1.2,0.76) -- (1.2,0.99);
  \node at (-0.35,0.5) {$L$};
  \node at (1.4,0.86) {$h$};
\end{tikzpicture}

\vspace{-2mm}
\item explicit time-stepping schemata:
{\footnotesize
\begin{align*}
\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
\end{align*}
}

@@

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$
\end{tabular}

\bigskip
\begin{itemize}
\item \alert{explicit} time-stepping \alert{implies many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] stress-balance solver scaling $\alpha$ is important, but even optimality ($\alpha=0$) cannot fix performance
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
    \item[$\circ$] parameterize cost of these solves as $O(m^{1+\beta})$
    \end{itemize}
\item nonetheless need $q$ model updates per year to integrate climate influences
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs), but in the \alert{non-local residual case}
    \begin{itemize}
    \item[$\circ$] this seems not to exist
    \item[$\circ$] the \alert{smoother} must compute an implicit-step SKE residual including surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ from a scalable Stokes solver
    \end{itemize}
\item near-optimal multilevel VI solvers exist for toy problems
    \begin{itemize}
    \item[$\circ$] Poisson equation obstacle problem (Brandt \& Cryer 1983; Gr\"aeser \& Kornhuber 2009)
    \end{itemize}
\item<2> join me on this quest?
\end{itemize}
\end{frame}
